\chapter{Theoretical Background}

\section{Terminology}

\subsection{Morphology}

The basic morphological unit is called a \textbf{word}. For the purpose of this thesis, the term \textit{word} is used in a strictly orthographic (spelling-based) sense, since the paper works with large data and any looser definition would be hard to apply consistently. \textcite{AndersonEtAl2022} suggest that a \textit{word} should be the smallest separable unit in a language. However, it is possible to break the word itself but it gets more difficult.

Smaller pieces of a word are called \textbf{morphemes}. The same morpheme can be spelled differently depending on the surrounding word parts (for example \textit{necessary} $\rightarrow$ \textit{necessari-ly}). This phenomenon is called \textbf{allomorphy}. The concrete physical realisations of morphemes are called \textbf{morphs}. This paper works with the orthographic form of words (it does not cluster related allomorphs into morphemes), so the main unit in the words are morphs.

A word can contain one or more morphemes. If there are multiple morphemes in a word, they can be divided into two categories: \textit{roots} and \textit{affixes}.

\textbf{Affixes} cannot stand on their own, they are always attached to a \textbf{base}. The smallest inseparable \textit{base} is called the \textbf{root} \parencite{AndersonEtAl2022}. Words can also contain multiple roots. This effect is called \textbf{compounding}. A criterion for identifying a second root is that the morpheme in question can independently form a word with its suffixes, in which it acts as the root. Examples include \textit{fire-fighter} or \textit{heart-break}. The set of all words derived from one root forms the root's derivational family.

% lepší než gap
When annotating data, a more robust definition of morph segmentation is necessary. If an annotator is unsure about a particular border, whether it splits two morphemes or lies inside a single morph, the following questions are suggested:
\begin{itemize}
    \item Does the language contain a related word that shares one of the two halves? This should identify the base.
    \item Does there exist a pattern of words sharing the other half but attached to different bases?
    \item Is the average native speaker able to separate the morphemes and feel that there is a border? Is a native speaker able to form a different word on the same base?
\end{itemize}


\newpage


Affixes can be further divided into the following categories:
\begin{itemize}
    \item A \textbf{prefix} is attached before the base.
    \item A \textbf{suffix} follows the base.
    \item A \textbf{circumfix} is attached around the base.
    \item An \textbf{infix} is inserted inside the base, and a \textbf{transfix} across the base.
    \item A \textbf{simultaneous affix} occurs at the same time as its base (in tonal and sign languages).
    \item An \textbf{interfix} ties together two bases, or two morphemes that are hard to pronounce in direct contact. \textit{(This affix is not mentioned by \textcite{AndersonEtAl2022}, but it is necessary for this thesis.)}
\end{itemize}

This thesis aims at basic morpheme identification, so some simplification was made. One or more roots can be identified in one word. All morphemes preceding the first root are called \textit{prefixes}, and all morphemes following the last root are called \textit{suffixes}. Any remaining morphemes are labelled \textit{interfixes}, which in this context simply denotes affixes that the language places between roots.

\newpage

\subsection{Corpus linguistics}

A corpus is a collection of texts. Its purpose can vary:
\begin{itemize}
    \item A \textbf{parallel} corpus serves to compare two or more languages. It contains the same texts translated into multiple languages.
    \item Corpora can be \textbf{annotated} with morphological and syntactic tags. One of the best resources of this kind is the Universal Dependencies project.
    \item \textbf{Frequency lists} are pre-aggregated data in which all tokens of a given word type are summed. The data are sorted in descending order of frequency.
\end{itemize}

Corpus data contain all words as they appear in the texts, which inherently means that they include inflected forms as well as typos, proper names, words from other languages, and other non-standard items. This property is not necessarily beneficial or harmful in itself. For example, some common typos can form a recognisable pattern.

Several other terms are commonly used in corpus terminology:

A \textbf{token} is a single unit of data. It is a word in the original form collected from the corpus. A word \textbf{type} is a category of identical tokens. Every type is unique and has a certain number of tokens (its occurrences in the corpus). The \textbf{token count} of the corpus represents the total amount of collected data, while the \textbf{type count} represents only the number of distinct words. Since this thesis works directly with frequency lists, the token count per word type is referred to as the \textit{frequency} of the word, and the total frequency in the corpus is an equivalent term for the token count.

A unit of words in a language can be abstracted as a \textbf{lexeme}. A lexeme is a set of word types that represent the same word in multiple \textbf{word-forms} defined by inflection. Word-forms do not change the inherent meaning of the lexeme, only its grammatical role. Inflection can be divided into the declination of nominals and the conjugation of verbs. A lexeme is usually present only once in a dictionary, under a basic form called a \textbf{lemma}. The other word-forms are typically predictable, or the dictionary lists the irregular forms, for example: \textit{child, children} . All forms of a lemma is called a \textbf{paradigm}. \parencite{Panocova2021}
% paradigm


\textbf{Hapax legomena} are word types that occur only once in the entire corpus.

\newpage

\subsection{Scripts}
\label{sec:scripts}

Due to the variety of languages covered in this thesis, a wide range of scripts is also present. This section introduces a basic typology of scripts to be used in later phases of the thesis.

First, there is a distinction between the terms \textbf{script} and \textbf{writing system}. A writing system is specific to each language and describes how the language uses a script, while a script is strictly the graphic form of the characters. \parencite{coulmas2003writing}


\subsubsection{Basic typology}

The typology of scripts is fundamental for studying complexity based on the written form of a language. Different scripts have different mechanisms that influence complexity measures. The typology was obtained from \textcite{coulmas2003writing}.

\begin{itemize}
    \item \textbf{Logographic}: the basic unit of the script is a word or a morpheme. Examples include Sumerian cuneiform and Chinese characters.
    \item \textbf{Syllabic}: the basic unit of the script is a syllable. Examples include Japanese kana and Cherokee.
    \item \textbf{Alphabetic}: the basic unit of the script is a phoneme and it is further divided into subtypes:
    \begin{itemize}
        \item \textbf{Abjad} scripts primarily represent consonants, while vowels are optional and indicated by diacritics. Examples include Arabic and Hebrew. The implication for complexity is that some words may be written identically, while having differences in pronunciation and meaning.
        
\begin{tabular}{lll}
\textarabic{كَتَبَ} & kataba & = he wrote \\
\textarabic{كُتِبَ} & kutiba & = it was written \\
\textarabic{كُتُب}  & kutub  & = books \\
\end{tabular}
        \item \textbf{Abugida} scripts represent vowels through marks attached to a consonant character. This group includes the Brāhmī-derived scripts such as Devanagari, Tamil and Thai, as well as Ethiopic. It is important to note that in UTF-8 these characters are usually represented as two independent characters. For example, the symbol read as \ttt{ki} --- `\textsanskrit{कि}' --- is a combination of \ttt{ka} --- `\textsanskrit{क}' and the vowel marker \ttt{i} --- `\textsanskrit{ि}', where the consonant comes first.
        \item  \textbf{Alphabet} scripts represent both consonants and vowels as independent characters. Examples are Greek-based scripts such as Latin and Cyrillic, as well as Armenian, Georgian and others. While Cyrillic usually represents one phoneme by one character, the Latin script has only 26 distinct letters, so alternations or combinations of letters are used to cover the full range of phonemes.
    \end{itemize}
\end{itemize}

Some scripts do not mark spaces between words, such as Chinese and Thai.

\newpage

\subsection{Morphological typology}
\label{chap:typology}

Different languages employ different morphological tools. Morphological typology is one of fundamental basics of linguistics, even though it is simplifying. The typology is presented here from the overview of \textcite{Panocova2021}. The typology below classifies a language by its dominant morphological strategy. It is important to note that this is an extreme simplification and many languages are not easily categorised, or display features of several categories. Nevertheless, it is useful for capturing basic relationships.

\subsection{Index of synthesis}

The index of synthesis describes how many morphemes are typically combined in a single word.

\begin{itemize}
    \item \textbf{Isolating (analytic)} --- languages with little morphology; grammatical meanings are expressed mainly by word order, particles, or auxiliary words. Examples: Chinese, Vietnamese, Thai, English.
    
    \item \textbf{Synthetic} --- languages where words often contain several morphemes expressing grammatical categories. Examples: Czech, Latin, Russian, Turkish, Finnish, Hungarian.
    
    \item \textbf{Polysynthetic} --- highly synthetic languages where one word may correspond to a whole sentence. Examples: Inuktitut, Greenlandic, Mohawk.
\end{itemize}

\subsubsection{Index of Fusion}

The index of fusion describes how clearly morphemes can be separated and how many meanings one morpheme expresses. It applies mainly to synthetic languages.

\begin{itemize}
    \item \textbf{Agglutinative} --- languages with clearly separable morphemes, each usually expressing one function. Examples: Turkish, Finnish, Hungarian, Japanese, Korean, Swahili.
    
    \item \textbf{Fusional} --- languages where one morpheme often expresses several grammatical meanings. Examples: Czech, Latin, Russian, Polish, Greek, German.
\end{itemize}

\subsubsection{Skalička's Typology}

Skalička further distinguishes two types of fusional morphology.

\begin{itemize}
    \item \textbf{Introflexive} --- languages using internal changes of roots or patterns. Examples: Arabic, Hebrew, Amharic.
    
    \item \textbf{Inflectional} --- languages with rich affixation, where one affix may express several categories at the same time. Examples: Czech, Latin, Ancient Greek, Russian, Sanskrit.
\end{itemize}

\subsubsection{Final typology}

In this work, following classification is used, combining the typologies described above. The categories are also listed in order of increasing complexity.
\begin{enumerate}
    \item \textbf{Isolating} (analytic)
    \item Synthetic --- \textbf{Agglutinative}
    \item Synthetic --- \textbf{Fusional} --- Inflectional
    \item Synthetic --- Fusional --- \textbf{Introflexive} (treated as a special case of fusional)
    \item \textbf{Polysynthetic}
\end{enumerate}



\newpage{}
\section{Morphological complexity}

Language complexity can be approached from two different perspectives. The first is called \textbf{integrative complexity} \parencite{Coltekin2022}. It measures how difficult it is for first-language learners (children acquiring their native language) and second-language learners to reach basic communication skills, or full proficiency, in the language. The two groups acquire a language in vastly different ways. Children learn without explicit rules: they just observe and repeat, and they are able to learn even the most difficult aspects of a language by heart, and only implicitly perceiving the underlying rules. The second group typically consists of adults or older children, who are already capable of understanding rules in their native language. This kind of measure can help decide which languages to learn first, or how much time and effort is needed to master a particular language.

%není lepší slovo?
The other approach is called \textbf{enumerative complexity} \parencite{Coltekin2022} and measures the complexity of a language from its structure. This method does not focus on the speakers learning the language, but on the language itself. The difficulty here is that it is not easy to determine which linguistic properties make a language more complex and which have the opposite effect. Some example methods are listed below. It is possible to
\begin{itemize}
    \item tie the measures to the previous approach --- which rules are \textbf{harder to learn};
    \item measure the \textbf{brain activity} of speakers while they use the language;
    \item measure how much of a conversation is \textbf{transmitted through a noisy environment} --- in other words, how well a listener can recover the meaning of sentences at varying levels of speech corruption;
    \item \textbf{compare different measures with one another} and check whether they capture the same underlying variable, that is, whether languages that score high on one measure tend to score high on others as well.
\end{itemize}

The implications of integrative complexity are obvious. The implications of enumerative complexity, on the other hand, may be more disguised. Language research helps us in general to understand how human communication works and how languages behave over time. It also helps us understand why certain changes occur independently in multiple languages. Is there a relationship between the simplification of a language and the number of its speakers, or the number of second-language learners?
The most fundamental question in the field of complexity is the equi-complexity hypothesis, which states that all languages possess roughly the same amount of complexity, tied to how much complexity humans need and can handle in everyday communication, but that this complexity manifests itself in different ways. For example, languages with little inflection tend to have a more rigid word order (compare English with Latin).

\begin{quote}
\textit{``Again, one can isolate the complexity of a language in phonemics, in morphophonemics, in tactics, etc.; but these isolable properties may hang together in such
a way that the total complexity of a language is approximately the same for all
languages.''} Rulon S. Wells 1954 \parencite{Newmeyer2022}
\end{quote}

In the footnote, Wells acknowledges that this idea comes from Charles Hockett (1952) \parencite{Newmeyer2022}.

\subsection{Enumerative complexity}

The earlier work by \textcite{Ehret2023} introduces a shared task featured at the ``Interactive Workshop on Measuring Language Complexity,'' which took place at the Freiburg Institute for Advanced Studies in September 2019, and lists several interesting approaches to measuring complexity. A few of the topics covered are, for example:
\begin{itemize}
    \item The complexity of Slavic languages correlates with the average altitude at which their speakers live. More remote Slavic languages tend to preserve their complexity.
    \item Syntagmatic redundancy, such as the use of multiple negatives, increases a language's absolute complexity, but paradoxically facilitates language acquisition and processing for its users.
    \item Languages used in a broader range of communicative contexts (with higher ``situational diversity'') tend to reduce their grammatical complexity over time in order to remain adaptable.
\end{itemize}

It also contains more general papers, one of which is by \textcite{Coltekin2022}. That study partially confirms, across multiple measures, that there is a correlation between them, suggesting an underlying phenomenon: a single shared notion of complexity. The concrete measures are listed in Chapter~\ref{sec:metrics}. The authors compare the ordering of languages based on various measures and the correlations between them, perform a PCA transformation to identify the main sources of variation, and also compare the results to the WALS (language feature) database.

\subsection{Position of this thesis}

The primary goal of this thesis is to investigate enumerative complexity in depth. To achieve this, various complexity metrics are tested on substantially larger corpus datasets covering a broader range of languages. Finally, these metrics are systematically compared with one another in order to evaluate their relationships and performance.
