
\chapter{Implementation}

 %TODO github

The whole computational work was done in Python programming language. The reasons for selecting it are the flexibility of the language, most of the programms were rewritten multiple times due to new findings; and the availability of libraries for data processing and visualization. The limitation of Python is its speed, which was a actually on a limit of practical use for the largest computation phases. Aggregation phase beeing the longest with multithreading taking more than one hour to complete. If more data will be necessary, it would be reasonable to switch to a faster language for the critical parts.

The graphs and results provided in the following chapters are generated via Jupyter notebooks.

The work was divided into two main stages: data processing and analysis.

The code base is divided into multiple folders. The main folders are \path{src/} for data processing python files, \path{notebooks/} for Jupyter notebooks, \path{data/}, \path{results/} and \path{test/}. All those folders are subsegmented into folders named after the phases of the work. 

\begin{itemize}
    \item \path{1_process/}
    \begin{itemize}
        \item \path{0_download/}
        \item \path{1_filter/}
        \item \path{3_truncate/}
    \end{itemize}
    \item \path{2_segmentation/} --- not implemented, reserved for future work
    \item \path{3_analysis/}
\end{itemize}

The data flow follows the same numbering. Each stage reads from \path{data/} or from the previous stage's output in \path{results/}, and outputs a clearly named CSV into \path{results/} folder. The numbers in folder names give the intended order of execution, so a new contributor can follow the pipeline just by walking through \path{src/} top to bottom.

\newpage

\section{Data collection}

Folder: \path{src/1_process/0_download/[name_of_the_source]/}

Data collection was either done manually directly from a webpage (Bible corpus, Universal Dependencies) or via scripts. The scripts are not part of any pipeline, since this is a one-time process and they are highly specific to the current setting of the receiving PC, the network and the data source. For the Leipzig corpus there is no API, so the script guesses the URL and downloading still requires some manual work to obtain the correct file names. For Glot500 there is an API. The program runs in two threads. One downloads batches of data and the other collects them and composes the per-language files. This separation keeps the network busy while the previous batch is still being parsed and aggregated into a frequency list.

Downloaded files are in the format \path{[language_code]_[corpus_name].csv}. They have two columns: word type and absolute token count, separated by a comma.

\section{Data processing}
\label{sec:impl_processing}

The data processing stage is operated via a makefile located in \path{src/1_process/}. The makefile wires together the underlying Python programs and provides a small set of common options. The python files themselves have many more possibilities and options resulting from the trial-and-error approach. The makefile exposes the following main targets:

\begin{lstlisting}
filter      # hapax-overlap and script checks -> ignored_files.csv
aggregate   # merge raw files per language, clean word types
statistics  # compute coverage tables for all languages
truncate    # cut each language at the chosen coverage
ppm         # add parts-per-million column to truncated files
all         # run the whole pipeline end to end
\end{lstlisting}

Common variables can be passed on the command line, for example \path{LANGS="ces eng"} to restrict the run to selected languages, \path{COVERAGE=94} to choose the truncation level, or \path{WORKERS=4} to set the number of parallel processes. A full description of every target and option is in an present markdown file (\path{src/1_process/preprocess.md}).

\subsection{Filtering}

Filtering does not touch the raw data. It only produces the file \path{results/1_process/1_filter/ignored_files.csv}, which lists raw files that should be skipped in later phases, together with the reason. The motivation is that filtering decisions are tuned several times during the project, and rerunning them must not destroy the original data. Two methods are currently in the pipeline: overlaping of the data, by hapax share between the files and a script checking. The methodology is described in the Section~\ref{sec:filtration}. Both checks are implemented as separate scripts and write their own report. The script \path{generate_ignore_list.py} merges these reports into the final ignore list. It is important to run the overlap check before the script check, and generate the ignored list after each phase, not to be influenced by some corpora being counted twice in finding the language dominant script.

\subsection{Aggregation}

 For each language, all non-ignored raw files are read, word types are normalised (boundary punctuation stripped, lowercased, entries without any letter dropped), and absolute frequencies are summed across files. The output is one CSV per language in \path{data/1_aggregated/}.

The aggregation step is the heaviest part of the pipeline. Basic multithreading would not help, since the memory is the actual bottleneck. The program needs to keep the whole frequency dictionary of the largest language in memory. There is a simple multithreading with one thread processing the largest wordlists and the another thread can process smaller wordlists in parallel, if it is provided with enough memory. Another way how to accelerate the process is not to generate a report. The report is important for seeing the normalization statistics, but it is not necessary to generate it every time.

For tokenization for languages which do not use spaces between words, the script \path{parse_words.py} is provided. It currently supports Chinese and Thai, which are provided with external Python libraries \path{jieba} and \path{pythainlp}. This is done in place on aggregated data. This script is automatically summing the frequencies.

Wordlists for languages are saved under this pattern: \path{[code of language]_[Code of script].csv}. The codes of languages are under the norm ISO~639-3 and they are three letters long. A norm ISO~15924 applies to the four letter script codes.

%language overview ?

\subsection{Truncation}

Truncation has two steps. The program \path{statistics.py} had to be run first. It generates a report, where is a row for every language and columns with token and word-type amounts for multiple truncating methods described in the Chapter~\ref{sec:truncation}. This report is referenced by notebooks generating graphs and the actual truncation algorithm. This process takes also significant amount of time. The truncation itself is on the other hand very fast. The program \path{truncate.py} simply reads the chosen coverage level column and copies the corresponding aggregated file up to the correct word-type count to \path{data/2_annotated/}. When several words share the cutoff frequency, the truncation samples words at random.

% rename annotated folder
\newpage
\section{Analysis}

Folder: \path{src/3_analyze/}.

The analysis stage is very simple: it iterates over the annotated or segmented files, computes a fixed set of metrics per language, and writes one row per language to a CSV. There are three independent producers.


\begin{itemize}
    \item \path{analyze.py} computes the corpus metrics (TTR, entropy, perplexity, Zipf slope, hapax counts, average word length) and writes \path{results.csv}.
    \item \path{morphs.py} does the same on the segmented data and writes \path{morphs.csv}, including the morph-level metrics and the morph-typology metrics (compounding index, affix deviation, affix entropy).
    \item \path{compressing/compress.py} computes the information in word structure metric.
\end{itemize}

The process can be runned by a makefile provided in the analyze folder with options corresponding to the phases.
%TODO implement the make

\begin{lstlisting}
analyze
morphs
compress
merge
\end{lstlisting}

All three producers share one small building block: a \path{Frequency} class that incrementally tracks total count, type count and the number of hapaxes, and exposes entropy and TTR as $O(1)$ queries. New metrics can therefore be added by writing a single function \path{f(freq) -> value} and registering it in the \path{METRICS} dictionary; the driver loop picks them up automatically.

The three CSVs are then joined together. \path{improve_results.py} merges \path{results.csv} with \path{language_overview.csv} (to add language names, other information, scripts and original corpus sizes) and with \path{compress.csv}, and produces \path{all_result.csv}. The subset (column-wise) of this file is presented in the appendix. This file is alsou the main source for the notebooks.

% one pipe for merging data

\section{Visualisation and reproducibility of figures}

Folder: \path{notebooks/}.

Every figure and table in this thesis is produced by a Jupyter notebook in \path{notebooks/1_process/} or \path{notebooks/3_analyze/}. To keep the figures synchronised with the text, a small helper module \path{notebooks/scripts/graph_export.py} is used. Inside a notebook, a figure is exported with one call:

\begin{lstlisting}
graph_export.save("ttrToEntropy", fig)
\end{lstlisting}

The helper saves the figure as a PDF, records in \path{graphs.yaml} which notebook produced which file together with a timestamp, and copies the result into \path{thesis/img/auto/} where it is picked up by the LaTeX build. A command-line mode (\path{python graph_export.py --update}) re-executes every registered notebook, so that all figures in the thesis can be regenerated from the current data with a single command. This was important in practice, since the data were rebuilt several times during the project.

\section{Tests}

Folder: \path{tests/}.

The processing and analysis programs are covered by \path{pytest} unit tests that mirror the \path{src/} folder structure. The tests focus on the most crucial parts of the pipeline to test whether the filtration, truncation methods and mainly the metrics calculations are correct. The shared fixture file \path{tests/1_process/conftest.py} provides synthetic frequency lists (two fake languages with controlled overlap patterns and script mixtures) used by several of these tests.

The processing tests (\path{tests/1_process/}) cover:
\begin{itemize}
    \item \path{test_filter.py} --- validates that the hapax-overlap detector correctly identifies duplicate corpora, that the script checker flags files written in an unexpected script, and that the ignore-list generator combines both reports correctly.
    \item \path{test_aggregate.py} --- verifies frequency summation across multiple source files, punctuation stripping, lowercasing, non-word removal, and the job scheduler that allocates large and small languages to parallel workers.
    \item \path{test_parse_words.py} --- tests the character segmentation for Chinese and Thai, including frequency conservation across segments and correct round-trip CSV I/O.
    \item \path{test_quality_check.py} --- checks percentile computation, dominant-script identification, and detection of encoding artifacts such as HTML entities.
    \item \path{test_statistics.py} --- validates coverage-level computation (how many word types cover $X$\% of tokens) against a small Zipfian distribution with a known answer.
    \item \path{test_truncate.py} --- ensures the truncation produces the exact type count dictated by the statistics table and the random sampling procedure.
\end{itemize}

The analysis tests (\path{tests/3_analyze/}) cover:
\begin{itemize}
    \item \path{test_analyze.py} --- tests the \path{Frequency} class and all corpus-level metrics (TTR, entropy, hapax ratio, average word length) against synthetic data with known mathematical properties.
    \item \path{test_morphs.py} --- validates morph-level metrics (morph TTR, morph entropy, perplexity, Zipf slope, frequency-weighted average length) and verifies that duplicate morpheme keys are merged correctly.
\end{itemize}

The whole suite is run from the project root with:

\begin{lstlisting}
pytest
\end{lstlisting}

A single phase or file can be selected by passing its path, for example \path{pytest} \path{tests/1_process}.



\section{Reusability}

The whole pipeline is designed so that adding a new data source or a new metric is a local change. A new corpus only needs a download script that produces files in the standard \path{[language_code]_[corpus_name].csv} format; the rest of the pipeline picks them up automatically. A new metric only needs one function and one entry in the corresponding \path{METRICS} dictionary. A new graph only needs one notebook cell and one call to \path{graph_export.save}. The intermediate CSVs are kept human-readable and small enough to inspect by hand, which made debugging across so many languages feasible without specialized tooling.